emmeans and varpred comparison

### 2021-10-08 compare_emmeans_varpred.rmd
why the difference: emmeans vs varpred

>>>>

\section{Why the differences?}

We actually don't know which method is correct. Let us consider a simple simulation:

- Outcome: hhsize
- Predictors:
	- Gender (40\% Males)
	- Age: Females are slightly older

```{r sim_compare, echo=TRUE}
## Simulation
N <- 100
extraAge <- 0.4
meanAge <- 0.5
sdAge <- 1
prop <- c(0.6, 0.4)
betas <- c(1.5, 0.5, 0.5)
gender <- sample(c("Female", "Male"), N, replace = TRUE, prob = prop)
df <- (data.frame(gender=gender)
	%>% mutate(muAge=ifelse(gender=="Female", meanAge+extraAge, meanAge)
		, age=rnorm(N, muAge, sdAge)
	)
	%>% select(-muAge)
)

mm <- model.matrix(~gender+age, df)
df$hhsize <- rnorm(N, mean=as.vector(mm %*% betas), sd=1)

## Model
mod <- lm(hhsize~gender+age, data=df)
print(mod)
```

Let us take a look at the observed marginals:

```{r observed_margins, echo=TRUE}
observed_margins <- (df
	%>% group_by(gender)
	%>% summarise_all(mean)
)
observed_margins
```

- **varpred** constructs model matrix by averaging _age_ within the levels of _gender_. The population average is the weighted (by the observed proportions) average of these averages:

```{r varpred_mm, echo=TRUE}
varpred_pred <- varpred(mod, "gender", within.category=TRUE, returnall=TRUE)
varpred_mm <- varpred_pred$raw$model.matrix
print(varpred_mm)
```

Estimates:
```{r varpred_pred}
print(varpred_pred)
```

By hand calculation:

```{r varpred_hand, echo=TRUE}
varpred_mm %*% coef(mod)
```

- **emmeans** uses the population average and seems not apply the weights:

```{r emmeans_mm, echo=TRUE}
emmeans_grid <- ref_grid(mod)
emmeans_mm <- emmeans_grid@grid
print(emmeans_mm)
```

Estimates:
```{r emmeans_pred}
emmeans_pred <- emmeans(emmeans_grid, spec=~gender)
print(emmeans_pred)
```

By hand calculation:

```{r emmeans_hand, echo=TRUE}
emmeans_mm <- emmeans_pred@linfct
print(emmeans_mm)

as.matrix(emmeans_mm) %*% coef(mod)
```

Mathematically, in **varpred**

\begin{align}
\mathbb{E}(Y|X) =
\begin{cases}
(\beta_0 + \alpha) + \beta_1 \bar{{Age_{M}}},& \text{ if } gender = Male\\\\
\beta_0 + \beta_1 \bar{{Age_{F}}},&  \text{ if } gender = Female\
\end{cases}
\end{align}

**emmeans**

\begin{align}
\mathbb{E}(Y|X) =
\begin{cases}
(\beta_0 + \alpha) + \beta_1 \bar{{Age}},& \text{ if } gender = Male\\\\
\beta_0 + \beta_1 \bar{{Age}},&  \text{ if } gender = Female\
\end{cases}
\end{align}
